---
title: "Working with ADaM Metadata"
subtitle: "Understanding CDISC ADaM Metadata Structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with ADaM Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r setup}
library(mighty.metadata)
```

## Introduction to ADaM Metadata

Analysis Data Model (ADaM) is a CDISC standard that defines the structure and metadata for analysis datasets used in clinical trials. ADaM metadata is critical for ensuring dataset compliance, documentation, and traceability from source data to analysis results.

The `mighty.metadata` package provides tools for working with ADaM metadata, focusing on:

1. Transforming raw metadata into a structured format
2. Tracking relationships between datasets
3. Managing method descriptions and algorithms
4. Documenting the derivation of analysis variables

## ADaM Metadata Structure

The ADaM standard defines several structural elements that are represented in the `mighty.metadata` package:

### Dataset Classes

ADaM datasets are classified into several types:

* **Subject Level Analysis Dataset (ADSL)**: Contains one record per subject
* **Basic Data Structure (BDS)**: Contains analysis parameters with associated results
* **Occurrence Data Structure (OCCDS)**: Represents data about events that may occur multiple times

### Metadata Levels

ADaM metadata exists at multiple levels:

* **Dataset-level metadata**: Information about the dataset as a whole
* **Column-level metadata**: Information about individual variables
* **Value-level metadata**: Information about specific values of a variable

## Variable Origins

ADaM variables have different origins, which determine how they are processed:

### Predecessor Variables

These are variables that come directly from source domains with minimal or no transformation:

```{r eval=FALSE}
# Example of predecessor variable in mighty.metadata format
list(
  column = "USUBJID",
  label = "Unique Subject Identifier",
  type = "Char"
)

# Example of renamed predecessor variable
list(
  column = "TRTP",
  label = "Planned Treatment",
  type = "Char",
  source = "ADSL.TRT01P"
)
```

### Derived Variables

These are variables created through transformations or calculations:

```{r eval=FALSE}
# Example of derived variable
list(
  column = "AVAL",
  label = "Analysis Value",
  type = "Num",
  method = "Derived as follows:\nIf AEPAIN='Y' then AVAL=1;\nElse AVAL=0;"
)
```

### Assigned Variables

These are variables with values assigned based on specific rules:

```{r eval=FALSE}
# Example of assigned variable
list(
  column = "PARAMCD",
  label = "Parameter Code",
  type = "Char",
  method = "Assigned: PARAMCD is set to 'WEIGHT'"
)
```

## Building ADaM Metadata

The `mighty.metadata` package processes raw metadata components to build a structured representation:

```{r eval=FALSE}
# Input metadata should contain:
# - source_tables: Dataset-level information
# - source_columns: Column-level information
# - source_values: Value-level information

adam_metadata <- build_adam_metadata(metadata)
```

The resulting structure has several key components:

### Dataset Structure

```{r eval=FALSE}
# Example dataset structure
dataset_structure <- list(
  dataset = "ADAE",
  label = "Adverse Events Analysis Dataset",
  class = "BASIC DATA STRUCTURE",
  structure = list(
    list(record_key = "USUBJID"),
    list(sort_key = c("AESEQ")),
    list(timing = c("ASTDT", "AENDT"))
  )
)
```

### Column Metadata

```{r eval=FALSE}
# Example column metadata
column_metadata <- list(
  list(
    column = "USUBJID",
    label = "Unique Subject Identifier",
    type = "Char"
  ),
  list(
    column = "AVAL",
    label = "Analysis Value",
    type = "Num",
    method = "Derived from VS.VSSTRESN where VS.VSTESTCD='SYSBP'"
  )
)
```

### Value Metadata

```{r eval=FALSE}
# Example value metadata
value_metadata <- list(
  list(
    column = "PARAMCD",
    whereclause = "PARAMCD='HR'",
    label = "Heart Rate",
    xmlcodelist = "HR"
  ),
  list(
    column = "PARAMCD",
    whereclause = "PARAMCD='SBP'",
    label = "Systolic Blood Pressure",
    xmlcodelist = "SBP"
  )
)
```

## Domain References

A key feature of the `mighty.metadata` package is its ability to track references between domains:

```{r eval=FALSE}
# Example referenced domains
referenced_domains <- list(
  ADSL = c("USUBJID", "TRT01P", "AGE"),
  VS = c("VSTESTCD", "VSSTRESN")
)
```

Domain references are built in two ways:

1. **Direct References**: From explicit predecessor variables (e.g., `source = "ADSL.TRT01P"`)
2. **Method Text References**: From analyzing the text of method fields

```{r eval=FALSE}
# Extract domain references from method text
references <- extract_domain_references("The variable is derived using ADSL.SEX and ADVS.AVAL")
print(references)  # Returns c("ADSL", "ADVS")
```

## Example ADaM Datasets

Here are examples of common ADaM dataset structures:

### ADSL (Subject Level)

```
STUDYID | USUBJID | AGE | SEX | RACE | ARM | TRT01P | TRT01A
--------|---------|-----|-----|------|-----|--------|--------
STUDY1  | 001-001 | 65  | M   | WHITE| A   | DRUG   | DRUG
STUDY1  | 001-002 | 42  | F   | BLACK| B   | PLACEBO| PLACEBO
```

### ADVS (Vital Signs - BDS)

```
STUDYID | USUBJID | PARAMCD | PARAM        | AVAL | AVALU
--------|---------|---------|--------------|------|------
STUDY1  | 001-001 | HR      | Heart Rate   | 72   | beats/min
STUDY1  | 001-001 | SBP     | Systolic BP  | 120  | mmHg
STUDY1  | 001-002 | HR      | Heart Rate   | 68   | beats/min
```

### ADAE (Adverse Events - OCCDS)

```
STUDYID | USUBJID | AESEQ | AETERM        | AESEV   | ASTDT     | AENDT
--------|---------|-------|---------------|---------|-----------|----------
STUDY1  | 001-001 | 1     | Headache      | MILD    | 2020-01-05| 2020-01-07
STUDY1  | 001-001 | 2     | Nausea        | MODERATE| 2020-01-10| 2020-01-15
STUDY1  | 001-002 | 1     | Fatigue       | MILD    | 2020-01-08| 2020-01-10
```

## Conclusion

The `mighty.metadata` package provides a comprehensive framework for working with ADaM metadata. By structuring metadata according to CDISC standards and adding features like domain references tracking, it facilitates the creation, management, and documentation of clinical trial analysis datasets.

Using this package, you can:

1. Transform raw metadata into structured ADaM format
2. Track relationships between domains and variables
3. Document derivation algorithms clearly
4. Generate Define.xml-compatible metadata
5. Store metadata in version-controllable YAML format

This approach enhances transparency, reproducibility, and compliance in clinical trial data analysis.
