---
title: "mighty.metadata"
subtitle: "CDISC ADaM Metadata Management in YAML Format"
vignette: >
  %\VignetteIndexEntry{mighty.metadata}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(mighty.metadata)
```

## Introduction

The `mighty.metadata` package provides tools for working with CDISC ADaM metadata in YAML format. This vignette demonstrates how to use the package to:

1. Access standard CDISC libraries
2. Extract and transform metadata 
3. Build structured ADaM metadata
4. Generate YAML files
5. View metadata in HTML format

## Accessing Standard Libraries

The package provides functions to access CDISC standard libraries:

```{r eval=FALSE}
# Get the path to CDISC standards
standard_path()

# List available versions of a standard
master_lookup("adam")

# Read master metadata file (not run)
# metadata <- read_master("adam", filename = "cst_adam_metadata_file.xlsx")
```

## Building ADaM Metadata

Once you have source metadata, you can transform it into a structured ADaM metadata format:

```{r eval=FALSE}
# Assuming you have a metadata object with the required components:
# - source_tables
# - source_columns
# - source_values

# Transform raw metadata into structured ADaM metadata
adam_metadata <- build_adam_metadata(metadata)
```

The `build_adam_metadata` function processes different types of columns:

- **Predecessor variables**: Inherit metadata from parent datasets
- **Derived variables**: Include method fields that describe derivation algorithms
- **Assigned variables**: Include assignment information

## Working with Domain References

The package can analyze metadata to extract domain references:

```{r eval=FALSE}
# Extract references from method text
# references <- extract_domain_references("The variable is derived using ADSL.SEX and ADVS.AVAL")
# print(references)  # Should return c("ADSL", "ADVS")

# Build referenced domains for a dataset
# referenced_domains <- build_referenced_domains(adam_metadata, "ADAE")
```

## Writing to YAML Format

After building your metadata structure, you can write it to YAML files:

```{r eval=FALSE}
# Write a single domain's metadata to a YAML file
yaml_file <- write_adam_domain_yaml(
  domain_data = adam_metadata$ADAE,
  domain_name = "ADAE",
  output_dir = "metadata"
)
```

The YAML output has several advantages:

- Method fields are formatted as multiline strings for readability
- Long text is automatically wrapped at word boundaries
- NULL/NA fields are omitted for cleaner output

## Viewing Metadata in HTML

The package provides functions to view metadata in HTML format:

```{r eval=FALSE}
# View define metadata in HTML format
view_define_html(adam_metadata$ADAE)
```

## Complete Workflow Example

Here's a complete example workflow (not run):

```{r eval=FALSE}
# 1. Read master metadata
metadata <- read_master("adam", filename = "cst_adam_metadata_file.xlsx")

# 2. Build ADaM metadata structure
adam_metadata <- build_adam_metadata(metadata)

# 3. Output directory for YAML files
output_dir <- "metadata/adam"

# 4. Write all domains to YAML files
for (domain_name in names(adam_metadata)) {
  write_adam_domain_yaml(
    domain_data = adam_metadata[[domain_name]],
    domain_name = domain_name,
    output_dir = output_dir
  )
}

# 5. View a specific domain in HTML
view_define_html(adam_metadata$ADAE)
```

## Summary

The `mighty.metadata` package streamlines the process of working with CDISC ADaM metadata by:

1. Providing access to standard libraries
2. Transforming raw metadata into a clean, structured format
3. Tracking domain references
4. Generating well-formatted YAML files
5. Offering HTML visualization

This approach makes metadata more readable, version-controllable, and easier to maintain than traditional Excel-based approaches.
