---
title: "YAML Format for ADaM Metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{YAML Format for ADaM Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mighty.metadata)
```

## Introduction to YAML for Metadata

YAML (YAML Ain't Markup Language) is a human-readable data serialization format that is particularly well-suited for representing hierarchical metadata structures. The `mighty.metadata` package uses YAML as the primary storage format for ADaM metadata due to several key advantages:

1. **Human Readability**: YAML is easy to read and edit directly, unlike binary formats or Excel spreadsheets
2. **Version Control**: YAML files can be effectively tracked in version control systems like Git
3. **Hierarchical Structure**: YAML naturally represents the nested structure of metadata
4. **Multiline Text Support**: Method descriptions and algorithms are cleanly represented

## ADaM Metadata YAML Structure

The mighty.metadata package organizes ADaM metadata in a structured YAML format following a defined schema that can be found in `inst/schema/adam_schema.yaml`. The main components are:

### Dataset-level Metadata (table_metadata)

```yaml
table_metadata:
  table: ADAE
  label: "Adverse Events Analysis Dataset"
  class: "BASIC DATA STRUCTURE"
  subclass: "ADVERSE EVENT"
  structure: "One record per adverse event per subject"
  keys: [USUBJID, AESEQ]
  comment: "Contains all adverse events reported during the study"
```

### Column-level Metadata (column_metadata)

Column metadata is stored as a collection of objects under the `column_metadata` key:

```yaml
column_metadata:
  - column: STUDYID
    label: "Study Identifier"
    
  - column: USUBJID
    label: "Unique Subject Identifier"
    
  - column: TRTP
    label: "Planned Treatment"
    source: ADSL.TRT01P
    
  - column: AVAL
    label: "Analysis Value"
    xmlcodelist: "AVALC"
    method: |-
      Derived as follows:
      If AEPAIN='Y' then AVAL=1;
      Else AVAL=0;
```

### Value-level Metadata (value_metadata)

Value-level metadata (when needed) is stored under the `value_metadata` key:

```yaml
value_metadata:
  - column: PARAMCD
    whereclause: "PARAMCD='HR'"
    label: "Heart Rate"
    
  - column: PARAMCD
    whereclause: "PARAMCD='SBP'"
    label: "Systolic Blood Pressure"
    method: |-
      Selected from ADVS where VSTEST='Systolic Blood Pressure'
```

### Referenced Domains (referenced_domains)

The package automatically builds a `referenced_domains` section based on analysis of column sources and method text:

```yaml
referenced_domains:
  ADSL:
    domain: "ADSL"
    variables:
      - USUBJID
      - TRT01P
      - AGE
  DM:
    domain: "DM"
    variables:
      - SEX
      - RACE
```

## YAML Formatting Features

The `mighty.metadata` package implements several specific formatting features to enhance the readability and maintainability of the YAML metadata:

### Method Field Formatting

Method fields are treated as literal block scalar style (`|-`) to preserve formatting while ensuring clean YAML output:

```yaml
method: |-
  The parameter is derived as follows:
  
  IF VISIT='BASELINE' THEN DO;
    AVAL = BASE;
    DTYPE = 'COPY';
  END;
  ELSE DO;
    AVAL = VSSTRESN;
  END;
```

### Text Wrapping

Long text strings are automatically wrapped at word boundaries to improve readability:

```yaml
origindescription: |-
  This is a long description that will be automatically wrapped 
  at word boundaries to ensure that lines aren't excessively 
  long in the YAML file.
```

### Null Value Handling

The package automatically omits NULL or NA values from the YAML output to maintain a clean and focused structure:

```yaml
# Without NULL handling
column: AVAL
label: "Analysis Value"
type: "Num"
origin: ~
predecessor: ~
xmlcodelist: ~

# With NULL handling (final output)
column: AVAL
label: "Analysis Value"
type: "Num"
```

## Examples

Here's an example of a complete YAML metadata file for a simple ADaM dataset following the schema:

```yaml
table_metadata:
  table: ADVS
  label: "Vital Signs Analysis Dataset"
  class: "BASIC DATA STRUCTURE"
  structure: "One record per vital sign parameter, per timepoint, per subject"
  keys: [USUBJID, PARAMCD, AVISITN, ATPTN]
  
column_metadata:
  - column: STUDYID
    label: "Study Identifier"
    
  - column: USUBJID
    label: "Unique Subject Identifier"
    
  - column: PARAMCD
    label: "Parameter Code"
    
  - column: PARAM
    label: "Parameter"
    
  - column: AVAL
    label: "Analysis Value"
    method: |-
      Set to VSSTRESN from VS domain.
      
  - column: AVALU
    label: "Analysis Value Units"
    
value_metadata:
  - column: PARAMCD
    whereclause: "PARAMCD='HR'"
    label: "Heart Rate"
    xmlcodelist: "HR"
    
  - column: PARAMCD
    whereclause: "PARAMCD='SBP'"
    label: "Systolic Blood Pressure"
    xmlcodelist: "SBP"
    method: |-
      Selected from VS where VSTESTCD='SYSBP'
    
referenced_domains:
  VS:
    domain: "VS"
    variables:
      - USUBJID
      - VSSEQ
      - VSTESTCD
      - VSSTRESN
  ADSL:
    domain: "ADSL"
    variables:
      - USUBJID
      - TRT01P
```

## Conclusion

The YAML format used by the `mighty.metadata` package provides a clean, readable, and maintainable representation of ADaM metadata that addresses many of the limitations of traditional Excel-based approaches. By storing metadata in YAML:

1. Changes are easily tracked in version control
2. Metadata becomes more readable and self-documenting
3. Complex structures like algorithms are properly formatted
4. Files can be programmatically generated and validated

This approach aligns with modern software development practices while still maintaining compatibility with CDISC standards.
